A community-maintained standard library of population genetic simulations

### 2019-12-18 popSimManu.tex
Many grammar and other minor fixes. Commented out Methods for simulations not presented. Added IM reference. We need to clarify which simulations are and are not included in the manuscript.

\newcommand{\stdpopsim}{\texttt{stdpopsim}\xspace}
\newcommand{\dadi}{$\partial a \partial i$\xspace}
\newcommand{\MSMC}{\texttt{MSMC}\xspace}
\newcommand{\smcpp}{\texttt{smc++}\xspace}
\newcommand{\stairwayplot}{\texttt{stairwayplot}\xspace}
\newcommand{\fastsimcoal}{\texttt{fastsimcoal2}\xspace}
\newcommand{\tskit}{\texttt{tskit}\xspace}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.pdf}
\caption{\textbf{Parameters estimated using a multi-population human model}.
Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
A) Data were generated by simulating
replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
inferred in \cite{international2007second}. B) For \dadi and \fastsimcoal we show parameters inferred
by fitting the depicted IM model, which includes population sizes, migration rates, and a split
time between CEU and YRI samples. C) Population size estimates for each population (rows)
from \dadi, \texttt{fastsimcoal2}, and \smcpp (columns). In shades of blue we show the estimated
% RNG: Updated caption to reflect what is shown.
$N(t)$ trajectories for each replicate. In black we show the census population size from the simulated model.
The population split date, $T_{div}$, is shown at
the bottom, with a common X-axis to the population size panels.}
\label{fig:IM_popn_human}
\end{center}
\end{figure}

\paragraph*{Multi-population demographic models}
As \texttt{stdpopsim} implements multi-population demographic models, we also
explored two population inference with several popular methods. Specifically,
we performed simulations under multi-population models for humans and \emph{Drosophila}
and inferred demographic parameters using \dadi, \fastsimcoal, and \smcpp.
For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
with constant population sizes and bi-directional migration \citep{hey2004im}. Our motivation for using
an IM model was to mimic the approach often used on empirical datasets, where a relatively
simple model is fit that may not reflect the true underlying demography.
For human models with more than two-populations (e.g. \cite{gutenkunst2009inferring})
this means that we are inferring parameters for a model which itself does
not match the model from which the data were generated (Figure
\ref{fig:IM_popn_human}A and B).

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the \cite{gutenkunst2009inferring} model. Our results highlight many
of the strengths and weaknesses of the different types of methods we used.
For instance, the SFS-based approaches where we fit simple IM models do not capture
recent exponential growth in the CEU population, however do consistently recover the
simulated YRI population size history. Moreover, these approaches allow for estimating
migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
of divergence times. By contrast, \smcpp is much better at capturing the recent exponential
growth in the CEU population, though the inferred population sizes are in general more
noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
estimates that are consistently underestimated.

Again, we can compare between species and look at the performance of these methods in
the context of a two-population model from \emph{Drosophila melanogaster}. Figure
\ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
the model of \cite{li2006inferring}, which includes
an ancestral population in Africa, then a population expansion with a population
split and bottleneck into a European population with no post-divergence migration.
Here again, we find that \dadi and \fastsimcoal are more consistent in their inferred
parameters, but ignore the brief population bottleneck in Europe. The inferred
demographic parameters from \smcpp are more noisy, though in some cases are better
able to capture the bottleneck in the Europe population.

Although these results do not represent an exhaustive benchmarking,
we have highlighted some of the strengths and weaknesses of these methods.
Future work should build on these results and undertake more in-depth comparisons
under a wider range of simulated demographic models.

@@

To conduct inference on these models, we applied three commonly used methods:
\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal~\citep{excoffier2013robust},
and \smcpp~\citep{terhorst2017robust}. As above, these methods were used
generally with default settings and we did not attempt to optimize their performance or fit
parameter-rich demographic models. As with the single population inference pipeline,
we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
our inference pipeline and maintain reproducibility.

For both \dadi and \fastsimcoal, we fit a two population
isolation-with-migration (IM) model, with constant population sizes, to the simulated
data from the HomSap and DroMel models. This IM model contains six parameters:
the ancestral population size, the size of population 1 after the split, the size of
population 2 after the split, the divergence time, and two migration rate parameters.
Importantly, this meant that for both species, the
fitted model did not match the simulated model (Figures \ref{fig:IM_popn_human} and \ref{fig:two_popn_fly}).
In the HomSap case, we therefore performed inference solely on the Africa
and Europe populations, meaning that the Asia population functioned as a ghost
population that was ignored by our inference. Our motivation for fitting this simple
IM model was to mimic the typical approach of two population inference on empirical
data, where the user is not aware of the `true' underlying demography and the inference
model is often misspecified. To ground-truth our inference approach, we also conducted
inference on a generic IM model that was identical to the model used for inference \ref{fig:generic_IM}.

@@

Following simulation, we outputted tree sequences and masked low-recombination
regions using the same approach described for the single population pipeline above. We
converted tree sequences into a two-dimensional site frequency spectrum for all
chromosomes in the appropriate format for \dadi and \fastsimcoal. For each simulation
replicate, we performed 10 runs of \dadi and \fastsimcoal and checked for convergence.
Detailed settings for \dadi and \fastsimcoal can be found in the Snakefile
on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
The parameter estimates from the top run of each simulation replicate
are shown in Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.pdf}
\caption{\textbf{Parameters estimated using a two-population \emph{Drosophila} model}.
Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp.
Data were generated by simulating
replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
inferred in \cite{comeron2012many}. See legend of Figure \ref{fig:IM_popn_human} for details.
In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history
as given by the census size for the simulated model.}
\label{fig:two_popn_fly}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/generic_IM.pdf}
\caption{\textbf{Parameters estimated from a generic IM model} Here we show estimates of $N(t)$
inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp. Data were generated by simulating
under a generic IM model with a human genome and \cite{international2007second} genetic map.
In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history
as given by the census size for the simulated model.}
\label{fig:generic_IM}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/dromel_migration_rates.pdf}
\caption{\textbf{Migration rate parameters estimated under a two-population \emph{Drosophila} model}.
Here we show inferred migration rates from \dadi and \fastsimcoal.
Data were generated by simulating replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
inferred in \cite{comeron2012many}.}
\label{fig:dromel_mig_rates}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homsap_migration_rates.pdf}
\caption{\textbf{Migration rate estimates for the human Gutenkunst model.}
Here we show inferred migration rates from \dadi and \fastsimcoal.
Data were generated by simulating
replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
inferred in \cite{international2007second}.}
\label{fig:homsap_mig_rates}
\end{center}
\end{figure}

>>>>

\newcommand{\stdpopsim}{\texttt{stdpopsim}\xspace}
\newcommand{\dadi}{$\partial a \partial i$\xspace}
\newcommand{\MSMC}{\texttt{MSMC}\xspace}
\newcommand{\smcpp}{\texttt{smc++}\xspace}
\newcommand{\stairwayplot}{\texttt{stairwayplot}\xspace}
\newcommand{\fastsimcoal}{\texttt{fastsimcoal}\xspace}
\newcommand{\tskit}{\texttt{tskit}\xspace}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.png}
\caption{\textbf{Parameters estimated using a multi-population human model}.
Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
A) Data were generated by simulating
replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
inferred in \cite{international2007second}. B) For \dadi and \texttt{fastsimcoal2} we show parameters inferred
by fitting the depicted IM model, which includes population sizes, migration rates, and a split
time between CEU and YRI samples. C) Population size estimates for each population (rows)
from \dadi, \texttt{fastsimcoal2}, and \smcpp (columns). In shades of blue we show the estimated
% RNG: Updated caption to reflect what is shown.
$N(t)$ trajectories for each replicate. In black we show the census population size from the simulated model.
The population split date, $T_{div}$, is shown at
the bottom, with a common X-axis to the population size panels.}
\label{fig:IM_popn_human}
\end{center}
\end{figure}

\paragraph*{Multi-population demographic models}
As \texttt{stdpopsim} implements multi-population demographic models, we also
explored parameter estimation of population divergence parameters. In particular
we performed simulations under multi-population models for humans and \emph{Drosophila}
and inferred simulated parameters using \dadi, \fastsimcoal, and \smcpp.
For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
% RNG: Added appropriate Jody Hey citation
with constant population sizes and bi-directional migration \citep{hey2004im}. For human
models with more than two-populations, e.g. \cite{gutenkunst2009inferring},
this means that we are inferring parameters for a model which itself does
not match the model from which the data were generated (Figure
\ref{fig:IM_popn_human}A and B), however this is common practice in the field
when applying IM-type models.

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the \cite{gutenkunst2009inferring} model of human migration out of Africa.
Each of the methods compared has stronger
and weaker points to their accuracy across parameters. For instance the simple IM-style
analyses we have performed do not capture recent population size change in CEU, however
they do a good job of estimating the simpler simulated YRI population size history from this model.

Again we can compare between species and look at the performance of these methods in
the context of a two-population model from \emph{Drosophila melanogaster}. Figure
\ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
the model of \cite{li2006inferring}, which includes
an ancestral population in Africa, then a population expansion with a population
split and bottleneck into a European population. These simulations are generated
using the recombination map of \cite{comeron2012many}. Here methods differ dramatically
in the estimates, with \smcpp yielding quite noisy estimates of population size
towards the more ancient past and all methods failing to capture the very brief
population bottleneck in Europe, although the IM model of course only allows for
two epochs in the population size history.

Again the goal of these analyses are not to perform thorough benchmarking of
the methods at hand under a wide range of simulations, but instead to show how
the tools that we have built could be used to facilitate such comparisons.

@@

To conduct inference on these models, we applied three commonly used methods:
\dadi~\citep{gutenkunst2009inferring}, \fastsimcoal2~\citep{excoffier2013robust},
and \smcpp~\citep{terhorst2017robust}. As above, these methods were used
generally with default settings and we did not attempt to optimize their performance or fit
parameter-rich demographic models. As with the single population inference pipeline,
we developed a \texttt{snakemake}~\citep{koster2012snakemake} workflow to manage
our inference pipeline and maintain reproducibility.

For both \dadi and \fastsimcoal2, we fit a two population
isolation-with-migration (IM) model, with constant population sizes, to the simulated
data from the HomSap and DroMel models. This IM model contains six parameters:
the ancestral population size, the size of population 1 after the split, the size of
population 2 after the split, the divergence time, and two migration rate parameters.
Importantly, this meant that for both species, the
fitted model did not match the simulated model (Figures \ref{fig:IM_popn_human} and \ref{fig:two_popn_fly}).
In the HomSap case, we therefore performed inference solely on the Africa
and Europe populations, meaning that the Asia population functioned as a ghost
population that was ignored by our inference. Our motivation for fitting this simple
IM model was to mimic the typical approach of two population inference on empirical
data, where the user is not aware of the `true' underlying demography and the inference
model is often misspecified.
% RNG: Not in present manuscript.
%To ground-truth our inference approach, we also conducted inference on a generic IM model that was identical to the model used for inference.

@@

Following simulation, we outputted tree sequences and masked low-recombination
regions using the same approach described for the single population pipeline above. We
converted tree sequences into a two-dimensional site frequency spectrum for all
chromosomes in the appropriate format for \dadi and \fastsimcoal2. For each simulation
replicate, we performed 10 runs of \dadi and \fastsimcoal2 and checked for convergence.
Detailed settings for \dadi and \fastsimcoal2 can be found in the Snakefile
on the popgensims/analysis GitHub repository (\url{https://github.com/popgensims/analysis}).
The parameter estimates from the top run of each simulation replicate
are shown in Figures \ref{fig:IM_popn_human}C and \ref{fig:two_popn_fly}C.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/d_mel_two_popn_comp.png}
\caption{\textbf{Parameters estimated using a two-population \emph{Drosophila} model}.
Here we show estimates of $N(t)$ inferred using \dadi, \texttt{fastsimcoal2}, or \smcpp.
Data were generated by simulating
replicate \emph{Drosophila} genomes under the \cite{li2006inferring} model and using the genetic map
inferred in \cite{comeron2012many}. See legend of Figure \ref{fig:IM_popn_human} for details.
In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:two_popn_fly}
\end{center}
\end{figure}
